---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "100%")
```

# athanor

**Last Updated:** `r Sys.Date()`

<!-- badges: start -->
<!-- badges: end -->

The goal of `athanor` is to integrate gene expression (GEX) and B cell receptor (BCR) data to improve B cell subtyping. The package provides functions for data preprocessing, feature extraction, concatenation, weighted nearest neighbors, and plotting.

## Installation

You can install the development version of `athanor` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("eba28/athanor")
```

## Example

```{r example}
library(athanor)

## basic example code
```

## Details

- `automated_annotation()` used to support Azimuth, but Azimuth is not currently available on CRAN and causes installation issues due to a hidden requirement of a Signac function that is no longer a part of Signac (`RunChromVAR()`)
- The functions were designed to operate on v5 of Seurat. If you are using an older version, you may have to change `layer` to `slot` in some of the code.
- Since the focus is on having multiple modalities, the GEX object created by `seurat_pipeline()` will have "rpca" and "rna.umap" as reductions instead of the usual "pca" and "umap" respectively.
