---
title: "Color palettes in athanor"
output:
  rmarkdown::html_vignette:
    code_folding: show
    df_print: kable
    fig_caption: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Color palettes in athanor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 300, fig.align = "center",
                      out.width = "100%")
```

`athanor` includes `named_colors`, a curated list of named character vectors
that map categorical values to hexadecimal colors. These can be passed as the
`clrs_specific` argument in plotting functions and are designed to be
consistent and colorblind-friendly.

In addition, a companion function called `plot_color_scale()`, generates a data-anchored
diverging color scale for continuous data which is particularly useful with Seurat's
`DotPlot`.

---

## Setup

```{r libs}
library(athanor)
library(dplyr)
library(ggplot2)
library(Matrix)
library(patchwork)
library(Seurat)

# helper to display a named color vector as a swatch strip
# could also use colorspace::swatchplot() or scales::show_col()
show_palette <- function(colors, title = "") {
  if (names(colors) %>% is.null()) {
    names(colors) <- seq_along(colors)
  }

  df <- data.frame(x = seq_along(colors),
                   label = factor(names(colors), levels = names(colors)),
                   fill = unname(colors))

  ggplot(df, aes(x = label, y = 1, fill = I(fill))) +
    geom_tile(color = "white", linewidth = 0.3) +
    labs(title = title) +
    theme_void() +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_text(size = 7, angle = 45, hjust = 1))
}
```

---

## 1. Overview of `named_colors`

Most of the names will match metadata columns in the Seurat objects and combined BCR data.

```{r overview}
# all palette categories available
sort(names(named_colors))
```

---

## 2. BCR biology palettes

### Isotypes and constant-region calls

```{r isotype-palette}
#| fig.dim = c(7, 2)

show_palette(named_colors$isotype, "Isotypes")
```

```{r c-call-palette}
#| fig.dim = c(10, 2)

show_palette(named_colors$c_call, "Constant-region calls")
```

### Isotype switching stage

```{r isotopestage-palette}
#| fig.dim = c(4, 2)

show_palette(named_colors$isotype_stage, "Isotype switching stage")
```

### Light chain locus

```{r light-palette}
#| fig.dim = c(4, 2)

show_palette(named_colors$light, "Light chain loci")
```

### Somatic hypermutation frequency bins

Note that the default binned colors are not named as the maximum bin depends on the dataset.

```{r mu-freq-palettes}
#| fig.dim = c(8, 5)

p1 <- show_palette(named_colors$mu_freq_bins_binary, "Binary bins (2)")
p2 <- show_palette(named_colors$mu_freq_bins_fewer, "3-bin scheme")
p3 <- show_palette(named_colors$mu_freq_bins, "5-bin scheme")
p4 <- show_palette(named_colors$mu_freq_iso, "SHM × isotype stage")

(p1 + p2) / (p3 + p4)
```

### V gene families

```{r vfamily-palettes}
#| fig.dim = c(9, 5)

p_igh <- show_palette(named_colors$v_call_igh_family, "IGHV families")
p_igk <- show_palette(named_colors$v_call_igk_family, "IGKV families")
p_igl <- show_palette(named_colors$v_call_igl_family, "IGLV families")

p_igh / p_igk / p_igl
```

---

## 3. General palettes

### Data types and modalities

```{r datatype-palette}
#| fig.dim = c(7, 2)

show_palette(named_colors$datatype, "Data types")
```

### Cell type annotations

```{r celltypes-palette}
#| fig.dim = c(10, 3)

show_palette(named_colors$cell_types_celltypist, "CellTypist cell types")
```

### CDR3 length

```{r cdr3-palette}
#| fig.dim = c(10, 2)

# show every 4th name to avoid overplotting
cdr3_sub <- named_colors$cdr3[seq(1, length(named_colors$cdr3), 4)]
show_palette(cdr3_sub, "CDR3 length (subset of range 4–41)")
```

### Embedding methods

```{r embeddings-palette}
#| fig.dim = c(8, 2)

show_palette(named_colors$embeddings, "Embedding methods")
```

---

## 4. Using palettes with plotting functions

All `plot_*` functions in `athanor` accept a `clrs_specific` argument. Pass a
`named_colors` entry whose names match the values in the metadata column being
plotted.

### Simulated Seurat object

```{r sim-obj}
n_cells <- 300
n_genes <- 200

obj <- sim_gex_splatter(num_genes = n_genes, num_cells = n_cells,
                        splatter_groups = c(0.5, 0.4, 0.1),
                        splatter_method = "groups")

obj <- seurat_pipeline(obj, nfeatures_RNA = 0, perc_mt = 100,
                       num_features = n_genes, num_pcs = 15, num_dims = 10,
                       k_param = 20, cluster_res = 0.5, verbose = FALSE)

# add BCR metadata
obj$isotype <- factor(sample(c("IgM", "IgD", "IgA", "IgG"), n_cells,
                              replace = TRUE, prob = c(0.35, 0.15, 0.30, 0.20)))
obj$mu_freq <- round(runif(n_cells, 0, 0.25), 3)
obj <- bin_mu_freq(obj)
```

```{r dimplot-isotype}
#| fig.dim = c(7, 5)

plot_dimplot(seurat_obj = obj, data_source = "Simulated",
             clrs_specific = named_colors$isotype, meta_col = "isotype",
             reduc = "rna.umap", title = "Isotype", legend_label = "Isotype",
             plot_label = FALSE)
```

```{r vln-mufreq}
#| fig.dim = c(9, 4)

plot_vln_feat(seurat_obj = obj,
              clrs_specific = named_colors$mu_freq_bins_fewer,
              feature = "mu_freq", meta_col = "mu_freq_bins_fewer",
              title = "SHMfrequency", reduc = "rna.umap")
```

If a value in your data has no matching entry in `clrs_specific`, `athanor`
will fall back to a default ggplot2 color scale. To add colors for custom
categories, extend the relevant palette:

```{r extend-palette}
my_colors <- c(named_colors$isotype, "NA" = "#cccccc")
my_colors
```

---

## 5. Continuous color scales with `plot_color_scale()`

`plot_color_scale()` anchors a diverging palette so that zero maps to white
and the positive / negative extremes are symmetric. This is particularly
useful for Seurat `DotPlot` output, where the default color mapping can
misrepresent where zero falls.

```{r color-scale-ggplot}
#| fig.dim = c(6, 4)

# continuous example with a regular ggplot2
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = mpg)) +
      geom_point(size = 3) +
      labs(title = "mtcars: mpg", subtitle = "Default scale")

# you can specify which column to pull values from
p_scaled <- plot_color_scale(plot = p, val_col = "mpg") +
              labs(title = "mtcars: mpg", subtitle = "Custom scale")

(p + p_scaled) & athanor::theme_bw_custom & athanor::labels_standard
```

```{r color-scale-dotplot}
#| fig.dim = c(10, 4)

# typical usage with a Seurat DotPlot
p <- DotPlot(obj, features = stringr::str_c("Gene", 1:10), assay = "RNA")

# you can pass a plot in with a pipe
p %>% plot_color_scale() # default parameters

# can change the colors as desired
plot_color_scale(plot = p, palette = rev(pals::brewer.brbg(n = 5)))
```

The `palette` argument accepts any vector of colors, defaulting to a
red–blue diverging scheme from `pals::brewer.rdbu`. Use `fill_by = "fill"` for
geoms that use the fill aesthetic rather than color.
