---
title: "Main processing pipeline: GEX, BCR, concatenation, and WNN"
output:
  rmarkdown::html_vignette:
    code_folding: show
    df_print: kable
    fig_caption: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Main processing pipeline: GEX, BCR, concatenation, and WNN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 300, fig.align = "center",
                      out.width = "100%")
```

This vignette walks through the full `athanor` pipeline on simulated data:

1. **GEX processing**: `seurat_pipeline()` + `add_annotations()`
2. **BCR metadata**: adding AIRR-derived features to the Seurat object
3. **WNN**: `run_wnn()` integrates GEX and BCR embeddings via Seurat's Weighted Nearest
   Neighbors
4. **Concatenation**: `concatenate_gex_bcr()` combines gene expression and BCR
   embeddings or features into a single assay
5. **Evaluation**: ADT-based metrics across methods

---

## Setup

```{r libs}
# TODO: switch to using real data

library(athanor)
library(dplyr)
library(ggplot2)
library(Matrix)
library(patchwork)
library(recipes)
library(Seurat)

set.seed(42)
```

---

## 1. GEX processing

### Simulate and build the initial object

`sim_gex_manual()` generates a sparse count matrix and wraps it in a Seurat
object. It also adds a `cell_id` column to the metadata, which is required
downstream by `run_wnn()`.

We also attach a small ADT assay (in real experiments this would hold
cell surface protein counts) since its presence is expected by `run_wnn()`.

```{r sim-gex}
num_cells <- 2000
num_genes <- 500
num_proteins <- 6
num_pcs <- 50
num_dims <- 20
k <- 20
data_desc <- "Simulated"

# TODO: put the memory and naive B cells into one cluster
obj <- sim_gex_splatter(num_genes = num_genes, num_cells = num_cells,
                        splatter_groups = c(0.4, 0.5, 0.1),
                        splatter_method = "groups")

# add a small ADT assay
adt_counts <- Matrix(as.integer(rexp(num_proteins * num_cells, rate = 0.3)),
                     nrow = num_proteins, ncol = num_cells, sparse = TRUE)
rownames(adt_counts) <- c("CD19", "CD21", "CD27", "CD38", "IGD", "IGM")
colnames(adt_counts) <- Cells(obj)
obj[["ADT"]] <- CreateAssay5Object(counts = adt_counts)

# confirm that cell_id is present
head(obj$cell_id)
```

### Run the GEX pipeline

`seurat_pipeline()` runs normalization, variable feature selection, scaling,
PCA, neighbor finding, and UMAP. Passing `nfeatures_RNA = 0` and
`perc_mt = 100` skips QC filtering (appropriate for simulated data with no
mitochondrial genes). A higher `cluster_res` gives more granular clusters,
which is useful for annotation.

```{r gex-pipeline}
obj <- seurat_pipeline(seurat_obj = obj, nfeatures_RNA = 0, perc_mt = 100,
                       num_features = num_genes, num_pcs = num_pcs,
                       num_dims = num_dims, k_param = k, cluster_res = 0.5,
                       verbose = FALSE)

# reductions and cluster column produced
names(obj@reductions)
levels(obj$seurat_clusters)
```

```{r gex-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj, plot_title = "GEX", data_source = data_desc,
             meta_col = "seurat_clusters", reduc = "rna.umap")
```

### Annotate clusters

`add_annotations()` maps a data frame of cluster-to-cell-type assignments onto
the Seurat object. The data frame must have one row per cluster in the same
order as `levels(seurat_clusters)`.

```{r annotate}
# TODO: annotate on a cell level instead to simulate automated annotation

n_clusters <- nlevels(obj$seurat_clusters)

# one cell type label per cluster
# cell_types <- sample(c("Memory B cells", "Naive B cells", "Plasma cells",
#                        "Transitional B cells"), size = n_clusters,
#                      replace = TRUE, prob = c(0.4, 0.4, 0.1, 0.1))

# we chose to have 3 clusters during simulation, so let's define them directly
cell_types <- c("Memory B cells", "Naive B cells", "Plasma cells")

annotations_df <- data.frame(CellType = cell_types)

obj <- add_annotations(seurat_obj = obj, annotations_df = annotations_df,
                       cell_types_col = "CellType",
                       clusters_col = "seurat_clusters",
                       annotations_col = "annotated_clusters")

table(obj$annotated_clusters)
```

```{r annotated-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "GEX", reduc = "rna.umap",
             meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

---

## 2. BCR data

### Embeddings

We'll quickly simulate some BCR embeddings.

```{r}
embeddings <- sim_airr_manual(num_cells = num_cells, num_dims = 256,
                              separator = "-")
colnames(embeddings) <- obj$cell_id  # must match cell_id

# examine the results
dim(embeddings) # dimensions by cells
embeddings[1:5, 1:5]
```

Let's make a Seurat object out of the embeddings:

```{r}
embeddings_obj <- bcr_embeddings_pipeline(embeddings = embeddings,
                                          embedding_type = "Manual",
                                          num_pcs = num_pcs,
                                          num_dims = num_dims, k_param = k,
                                          verbose = TRUE)
```

```{r bcr-umap}
#| fig.dim = c(8, 6)

# just for plotting (since we set cells to be the same here)
embeddings_obj$annotated_clusters <- obj$annotated_clusters

plot_dimplot(seurat_obj = embeddings_obj, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "BCR Embeddings",
             reduc = "bcr.umap", meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

### Metadata

In a real experiment these columns come from `gex_add_airr()` applied to an
AIRR-formatted table. Here we simulate them directly so that the downstream
steps are self-contained.

- `cdr3_aa_length` = CDR3 length bucket (ordered factor)
  - Can also be numeric
- `isotype` = BCR isotype (categorical)
- `mu_freq` = somatic hypermutation frequency (numeric)

```{r bcr-meta}
obj$cdr3_aa_length <- factor(sample(c("Short", "Medium", "Long"), num_cells,
                                    replace = TRUE),
                             levels = c("Short", "Medium", "Long"),
                             ordered = TRUE)
obj$isotype <- factor(sample(c("IgM", "IgD", "IgG", "IgA"), num_cells,
                             replace = TRUE,
                             prob = c(0.35, 0.15, 0.35, 0.15)))
obj$mu_freq <- round(runif(num_cells, min = 0, max = 0.3), 3)

obj[[]] %>%
  select(mu_freq, cdr3_aa_length, isotype) %>%
  summary()

# let's say that this object's cells all have paired BCRs too
obj$Has_BCR <- TRUE
obj$paired_light <- TRUE # TODO: test this
```

---

## 3. WNN (Embeddings)

`run_wnn()` integrates two modalities via Weighted Nearest Neighbors. It
expects:

- a Seurat object with a `cell_id` column in the metadata
- a BCR `embeddings` matrix (features × cells) whose column names match
  `cell_id`
  
You can technically also run this with BCR features but we do not recommend it.

```{r run-wnn}
obj_wnn <- run_wnn(seurat_obj = obj, embeddings = embeddings,
                   embedding_type = "Simulated",
                   pc_gex = 10, pc_bcr = 10, k_param = 20, cluster = TRUE,
                   cluster_res = list("GEX" = 0.5, "BCR" = 0.5, "WNN" = 0.5),
                   verbose = FALSE)

# bin the mutation frequency for plotting
obj_wnn <- bin_mu_freq(obj_wnn)

names(obj_wnn@reductions)
```

### Visualization

UMAPs:

```{r wnn-umap}
#| fig.dim = c(14, 5)

p_gex <- plot_dimplot(obj_wnn, plot_title = "GEX", data_source = data_desc,
                      clrs_specific = named_colors$cell_types_celltypist,
                      meta_col = "annotated_clusters", reduc = "rna.umap",
                      plot_label = FALSE, legend_label = "Cell Type")
p_bcr <- plot_dimplot(obj_wnn, plot_title = "BCR", data_source = data_desc,
                      clrs_specific = named_colors$cell_types_celltypist,
                      meta_col = "annotated_clusters", reduc = "bcr.umap",
                      plot_label = FALSE, legend_label = "Cell Type")
p_wnn <- plot_dimplot(obj_wnn, plot_title = "WNN", data_source = data_desc,
                      clrs_specific = named_colors$cell_types_celltypist,
                      meta_col = "annotated_clusters", reduc = "wnn.umap",
                      plot_label = FALSE, legend_label = "Cell Type")

# use patchwork
(p_gex + p_bcr + p_wnn) + plot_layout(nrow = 1, guides = "collect")
```

Modality weights:

```{r wnn-mw-box}
#| fig.dim = c(14, 5)

# by isotype
plot_mws(seurat_obj = obj_wnn, second_assay = "BCR",
         clrs_specific = named_colors$isotype, split_by = "isotype",
         facet_by = "annotated_clusters", y_axis_label = "Isotype")

# by SHM frequency
plot_mws(seurat_obj = obj_wnn, second_assay = "BCR",
         clrs_specific = named_colors$mu_freq_bins, split_by = "mu_freq_bins",
         facet_by = "annotated_clusters", y_axis_label = "SHM Frequency Bins")
```

`object_overview()` prints a brief description of the post-WNN object,
including assay sizes and the number of PCs used per modality.

```{r wnn-summary}
object_overview(obj_wnn)
```

---

## 4. Concatenation (Embeddings)

`concatenate_gex_bcr()` combines the RNA counts with processed BCR embeddings into a new `RNA_BCR` assay, then runs PCA and UMAP on the joint space.

* `stage = "raw"` means that  GEX genes are concatenated with the BCR embeddings
* `stage = "reduced_gex"` means that GEX PCs are concatenated with BCR embeddings
* `stage = "reduced_both"` means that GEX PCs are concatenated with BCR PCs

```{r concatenate-embeddings}
# TODO: remove this step
embeddings_obj$Has_BCR <- TRUE
embeddings_obj$paired_light <- TRUE

# merge the BCR embeddings into the GEX obj
gex_bcr_obj <- merge_gex_bcr(gex_obj = obj, bcr_obj = embeddings_obj,
                             transfer_reductions = TRUE, verbose = TRUE)

# keep a copy so run_wnn() below starts from the same object
obj_cat <- concatenate_gex_bcr(seurat_obj = gex_bcr_obj, stage = "reduced_both",
                               input_type = "embeddings",
                               gex_reduction = "rpca", num_dims = num_dims)

# new assay and reductions
names(obj_cat@assays)
names(obj_cat@reductions)
```

```{r concatenated-embeddings-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj_cat, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "Concatenated GEX & BCR Embeddings",
             reduc = "rna_bcr.umap", meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

```{r concatenated_embeddings-summary}
object_overview(obj_cat)
```

---

## 5. Concatenation (Features)

`concatenate_gex_bcr()` combines the RNA counts with processed BCR feature
columns into a new `RNA_BCR` assay, then runs PCA and UMAP on the joint space.

* `stage = "raw"` means that  GEX genes are concatenated with the BCR features
* `stage = "reduced_gex"` means that GEX PCs are concatenated with BCR features
* `stage = "reduced_both"` means that GEX PCs are concatenated with BCR PCs
  * This is not recommended as BCR features already tends to be a short list, so PCs are not informative

`cols_to_include` lists the BCR metadata columns to pull into the assay.
Internally, `process_bcr_features()` normalizes numeric columns, converts
ordered factors to ordinal scores, and one-hot encodes categoricals.

```{r concatenate-features}
# keep a copy so run_wnn() below starts from the same object
obj_cat <- concatenate_gex_bcr(seurat_obj = obj, stage = "reduced_gex",
                               input_type = "features",
                               cols_to_include =
                                 c("cdr3_aa_length", "isotype", "mu_freq"),
                               num_dims = num_dims)

# new assay and reductions
names(obj_cat@assays)
names(obj_cat@reductions)
```

```{r concatenated-features-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj_cat, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "Concatenated GEX & BCR Features", reduc = "rna_bcr.umap",
             meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

```{r concatenated-features-summary}
object_overview(obj_cat)
```

---

## 6. Evaluation

---

## Summary of the pipeline

| Step | Function | Key output(s) |
|------|----------|-----------|
| GEX simulation | `sim_gex_manual()` | Seurat object with `cell_id` |
| GEX pipeline | `seurat_pipeline()` | `RNA` assay, `rpca`, `rna.umap`, `seurat_clusters` |
| BCR simulation | `sim_bcr_manual()` | Matrix of embedding values |
| BCR pipeline | `bcr_embeddings_pipeline()` | `BCR` assay, `bpca`, `bcr.umap` |
| Annotation | `add_annotations()` | `annotated_clusters` in metadata |
| Addition of BCR features | `gex_add_airr()` | BCR columns in object metadata |
| Weighted nearest neighbors | `run_wnn()` | `wnn.umap` |
| Concatenation | `concatenate_gex_bcr()` | `RNA_BCR` assay, `rna_bcr.pca`, `rna_bcr.umap` |
| Description | `object_overview()` | printed summary |
