---
title: "Main processing pipeline: GEX, concatenation, and WNN"
output:
  rmarkdown::html_vignette:
    code_folding: show
    df_print: kable
    fig_caption: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Main processing pipeline: GEX, concatenation, and WNN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 300, fig.align = "center",
                      out.width = "100%")
```

This vignette walks through the full `athanor` pipeline on simulated data:

1. **GEX processing**: `seurat_pipeline()` + `add_annotations()`
2. **BCR metadata**: adding AIRR-derived features to the Seurat object
3. **Concatenation**: `concatenate_gex_bcr()` combines gene expression and BCR
   features into a single assay
4. **WNN**: `run_wnn()` integrates GEX and BCR embeddings via Seurat's Weighted Nearest
   Neighbors
5. **Evaluation**: ADT-based metrics across methods

---

## Setup

```{r libs}
library(athanor)
library(dplyr)
library(ggplot2)
library(Matrix)
library(patchwork)
library(recipes)
library(Seurat)

set.seed(42)
```

---

## 1. GEX processing

### Simulate and build the initial object

`sim_gex_manual()` generates a sparse count matrix and wraps it in a Seurat
object. It also adds a `cell_id` column to the metadata, which is required
downstream by `run_wnn()`.

We also attach a small ADT assay (in real experiments this would hold
cell surface protein counts) since its presence is expected by `run_wnn()`.

```{r sim-gex}
num_cells <- 2000
num_genes <- 500
num_proteins <- 6
data_desc <- "Simulated"

# TODO: put the memory and naive B cells into one cluster
obj <- sim_gex_splatter(num_genes = num_genes, num_cells = num_cells,
                        splatter_groups = c(0.2, 0.5, 0.3),
                        splatter_method = "groups")

# add a small ADT assay
adt_counts <- Matrix(as.integer(rexp(num_proteins * num_cells, rate = 0.3)),
                     nrow = num_proteins, ncol = num_cells, sparse = TRUE)
rownames(adt_counts) <- c("CD19", "CD21", "CD27", "CD38", "IGD", "IGM")
colnames(adt_counts) <- Cells(obj)
obj[["ADT"]] <- CreateAssayObject(counts = adt_counts)

# confirm that cell_id is present
head(obj$cell_id)
```

### Run the GEX pipeline

`seurat_pipeline()` runs normalization, variable feature selection, scaling,
PCA, neighbor finding, and UMAP. Passing `nfeatures_RNA = 0` and
`perc_mt = 100` skips QC filtering (appropriate for simulated data with no
mitochondrial genes). A higher `cluster_res` gives more granular clusters,
which is useful for annotation.

```{r gex-pipeline}
obj <- seurat_pipeline(obj, nfeatures_RNA = 0, perc_mt = 100,
                       num_features = num_genes, num_pcs = 15, num_dims = 10,
                       k_param = 20, cluster_res = 0.5, verbose = FALSE)

# reductions and cluster column produced
names(obj@reductions)
levels(obj$seurat_clusters)
```

```{r gex-umap}
#| fig.dim = c(8, 6)

plot_dimplot(obj, assay = "GEX", data_source = data_desc,
             meta_col = "seurat_clusters", reduc = "rna.umap")
```

### Annotate clusters

`add_annotations()` maps a data frame of cluster-to-cell-type assignments onto
the Seurat object. The data frame must have one row per cluster in the same
order as `levels(seurat_clusters)`.

```{r annotate}
n_clusters <- nlevels(obj$seurat_clusters)

# one cell type label per cluster
cell_types <- sample(
  c("Memory B cells", "Naive B cells", "Plasma cells", "Transitional B cells"),
  size = n_clusters, replace = TRUE
)

annotations_df <- data.frame(CellType = cell_types)

obj <- add_annotations(obj, annotations_df = annotations_df,
                       cell_types_col = "CellType", clusters_col = "seurat_clusters",
                       annotations_col = "annotated_clusters")

table(obj$annotated_clusters)
```

```{r annotated-umap}
#| fig.dim = c(8, 6)

plot_dimplot(obj, data_source = data_desc,
             assay = "GEX", reduc = "rna.umap",
             meta_col = "annotated_clusters", annotated = TRUE,
             plot_label = FALSE, legend_label = "Cell Type")
```

---

## 2. BCR metadata

In a real experiment these columns come from `gex_add_airr()` applied to an
AIRR-formatted table. Here we simulate them directly so that the downstream
steps are self-contained.

- `cdr3_aa_length` = CDR3 length bucket (ordered factor)
  - Can also be numeric
- `isotype` = BCR isotype (categorical)
- `mu_freq` = somatic hypermutation frequency (numeric)

```{r bcr-meta}
obj$cdr3_aa_length <- factor(sample(c("Short", "Medium", "Long"), num_cells,
                                    replace = TRUE),
                             levels = c("Short", "Medium", "Long"),
                             ordered = TRUE)
obj$isotype <- factor(sample(c("IgM", "IgD", "IgG", "IgA"), num_cells,
                             replace = TRUE,
                             prob = c(0.35, 0.15, 0.35, 0.15)))
obj$mu_freq <- round(runif(num_cells, min = 0, max = 0.3), 3)

obj[[]] %>%
  select(mu_freq, cdr3_aa_length, isotype) %>%
  summary()
```

---

## 3. Concatenation

`concatenate_gex_bcr()` combines the RNA counts with processed BCR feature
columns into a new `RNA_BCR` assay, then runs PCA and UMAP on the joint space.
`pca_stage = "Before"` means the BCR features are concatenated *before* PCA;
set it to `"After"` to combine pre-computed GEX and BCR PCA embeddings instead.

`cols_to_include` lists the BCR metadata columns to pull into the assay.
Internally, `process_bcr_features()` normalizes numeric columns, converts
ordered factors to ordinal scores, and one-hot encodes categoricals.

```{r concatenate}
# keep a copy so run_wnn() below starts from the same object
obj_cat <- concatenate_gex_bcr(obj, pca_stage = "Before",
                               cols_to_include =
                                 c("cdr3_aa_length", "isotype", "mu_freq"),
                               normalize = TRUE, num_dims = 10)

# new assay and reductions
names(obj_cat@assays)
names(obj_cat@reductions)
```

```{r concatenated-umap}
#| fig.dim = c(8, 6)

plot_dimplot(obj, data_source = data_desc,
             assay = "Concatenated GEX & BCR", reduc = "rna_bcr.umap",
             meta_col = "annotated_clusters", annotated = TRUE,
             plot_label = FALSE, legend_label = "Cell Type")
```

---

## 4. WNN

`run_wnn()` integrates two modalities via Weighted Nearest Neighbors. It
expects:

- a Seurat object with a `cell_id` column in the metadata
- a BCR `embeddings` matrix (features × cells) whose column names match
  `cell_id`

Here we simulate an immune2vec-style embedding matrix directly.

```{r bcr-embeddings}
num_dims_bcr <- 20

bcr_embeddings <-
  matrix(round(runif(num_cells * num_dims_bcr, min = -0.6, max = 0.6), 9),
         nrow = num_dims_bcr, ncol = num_cells)
rownames(bcr_embeddings) <- paste0("Dim-", seq_len(num_dims_bcr))
colnames(bcr_embeddings) <- obj$cell_id  # must match cell_id

# verify alignment
all(colnames(bcr_embeddings) == obj$cell_id)
```

```{r run-wnn}
obj_wnn <- run_wnn(obj, embeddings = bcr_embeddings,
                   embedding_type = "Simulated",
                   pc_gex = 10, pc_bcr = 10, k_param = 20, cluster = TRUE,
                   cluster_res = list("GEX" = 0.5, "BCR" = 0.5, "WNN" = 0.5),
                   show_output = FALSE)

names(obj_wnn@reductions)
```

### Visualization

```{r wnn-umap}
#| fig.dim = c(14, 5)

p_gex <- plot_dimplot(obj_wnn, assay = "GEX", data_source = data_desc,
                      meta_col = "annotated_clusters", reduc = "rna.umap",
                      plot_label = FALSE, legend_label = "Cell Type")
p_bcr <- plot_dimplot(obj_wnn, assay = "BCR", data_source = data_desc,
                      meta_col = "annotated_clusters", reduc = "bcr.umap",
                      plot_label = FALSE, legend_label = "Cell Type")
p_wnn <- plot_dimplot(obj_wnn, assay = "WNN", data_source = data_desc,
                      meta_col = "annotated_clusters", reduc = "wnn.umap",
                      plot_label = FALSE, legend_label = "Cell Type")

# use patchwork
(p_gex + p_bcr + p_wnn) + plot_layout(nrow = 1, guides = "collect")
```

### Object summary

`extract_wnn_vars()` prints a brief description of the post-WNN object,
including assay sizes and the number of PCs used per modality.

```{r wnn-summary}
extract_wnn_vars(obj_wnn, gex_pca = "rpca", other_pca = "bpca",
                 other_type = "BCR")
```

---

## 5. Evaluation

---

## Summary of the pipeline

| Step | Function | Key output |
|------|----------|-----------|
| Simulate GEX | `sim_gex_manual()` | Seurat object with `cell_id` |
| GEX pipeline | `seurat_pipeline()` | `rpca`, `rna.umap`, `seurat_clusters` |
| Annotate | `add_annotations()` | `annotated_clusters` in metadata |
| Add BCR features | `gex_add_airr()` / manual | BCR columns in metadata |
| Concatenation | `concatenate_gex_bcr()` | `RNA_BCR` assay, `rna_bcr.umap` |
| WNN | `run_wnn()` | `bpca`, `bcr.umap`, `wnn.umap` |
| Describe object | `extract_wnn_vars()` | printed summary |
