---
title: "Visualizing scRNA-Seq data with plot_dimplot()"
output:
  rmarkdown::html_vignette:
    # code_download: yes
    code_folding: show
    # css: ./"style.css"
    # dev: "png"
    df_print: kable
    fig_caption: yes
    # keep_md: no
    # self_contained: true
    # theme: yeti
    toc: yes
    toc_depth: 3
    # toc_float:
    #     collapsed: false
    #     smooth_scroll: false
vignette: >
  %\VignetteIndexEntry{Visualizing scRNA-Seq data with plot_dimplot()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 300, fig.align = "center",
                      out.width = "100%")
```

`plot_dimplot()` wraps Seurat's `DimPlot()` with consistent styling and a
flexible `meta_col` argument, making it easy to overlay any categorical metadata
column onto a UMAP (or other dimensionality reduction).  This vignette shows
three common use cases:

1. **GEX metadata**: Seurat clusters and cell-type annotations
2. **BCR metadata**: BCR features transferred from paired AIRR data
3. **ADT expression**: cell surface protein levels via `FeaturePlot()`

## Setup

```{r}
library(athanor)
library(ggplot2)
library(Matrix)
library(patchwork)
library(Seurat)

set.seed(42)
```

### Initial GEX object

```{r build-object}
# could change these
num_genes <- 400
num_cells <- 200
num_proteins <- 6
data_desc <- "Simulated"
named_clrs <- list() # so that there aren't a bunch of color variables

# TODO: use my simple simulator functions
gex_counts <- Matrix(as.integer(rexp(num_genes * num_cells, rate = 0.5)),
                     nrow = num_genes, ncol = num_cells, sparse = TRUE)
adt_counts <- Matrix(as.integer(rexp(num_proteins * num_cells, rate = 0.3)),
                     nrow = num_proteins, ncol = num_cells, sparse = TRUE)

rownames(gex_counts) <- paste0("Gene-", seq_len(num_genes))
colnames(gex_counts) <- paste0("Cell-", seq_len(num_cells))
rownames(adt_counts) <- c("CD19", "CD21", "CD27", "CD38", "IGD", "IGM")
colnames(adt_counts) <- colnames(gex_counts)

obj <- CreateSeuratObject(counts = gex_counts, project = "Example")
obj[["ADT"]] <- CreateAssay5Object(counts = adt_counts)

# TODO: show real QC
obj <- seurat_pipeline(obj, nfeatures_RNA = 0, perc_mt = 100,
                       num_features = 400, num_pcs = 15, num_dims = 10,
                       k_param = 15, cluster_res = 0.5, verbose = FALSE)

# add some B cell types
# TODO: include non-B cell types as well
obj$annotated_clusters <- sample(c("Memory B cells", "Naive B cells",
                                   "Plasma cells", "Transitional B cells"),
                                 num_cells, replace = TRUE)
```

### Adding BCR metadata

After running `gex_add_airr()` and `process_bcr_features()` on real data,
BCR-derived columns are available in `obj@meta.data`.  Here we simulate them
for illustration:

```{r bcr-metadata}
# TODO: correlate these values with cell types instead of having them be random

# TODO: change these to actual numbers
obj$cdr3_aa_length <-
  sample(c("Short", "Medium", "Long"), num_cells, replace = TRUE)
obj$isotype <-
  sample(c("IgM", "IgD", "IgG", "IgA", "IgE"), num_cells, replace = TRUE,
         prob = c(0.35, 0.15, 0.30, 0.15, 0.05))
obj$locus_light <-
  sample(c("IGK", "IGL"), num_cells, replace = TRUE)
obj$mu_freq_bins <-
  sample(c("0%", "0% to 1%", "1% to 5%", "5% to 10%", "10% to 25%"),
         num_cells, replace = TRUE)
obj$v_call_family <-
  sample(paste0("IGHV", 1:7), num_cells, replace = TRUE)
```

---

## 1. GEX Metadata

### Seurat clusters

The simplest call colors cells by `seurat_clusters` (the default `meta_col`).
Passing the assay name to `assay` populates the plot title.

```{r gex-clusters}
#| fig.dim = c(8, 7)

# could also just use umap
plot_dimplot(seurat_obj = obj, data_source = data_desc,
             title = "GEX", reduc = "rna.umap")
```

To suppress labels and the legend:

```{r gex-clusters-clean}
#| fig.dim = c(8, 7)

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             title = "GEX", reduc = "rna.umap",
             plot_label = FALSE, include_legend = FALSE)
```

### Cell type annotations

Sort the cell type levels alphabetically.

```{r gex-annotations}
#| fig.dim = c(8, 7)

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             title = "GEX", reduc = "rna.umap",
             meta_col = "annotated_clusters", plot_label = FALSE)
```

### Highlighting specific populations

Pass a character vector to `highlight` to draw a subset of cells on top
of a grey background.

```{r gex-highlight}
#| fig.dim = c(8, 7)

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             title = "GEX", reduc = "rna.umap",
             meta_col = "annotated_clusters",
             highlight = c("Naive B cells", "Plasma cells"))
```

---

## 2. BCR Metadata

BCR-derived columns from `process_bcr_features()` and `gex_add_airr()` are
transferred into the GEX Seurat object, so they are plotted exactly like any
other metadata column.

### CDR3 length

```{r bcr-cdr3}
#| fig.dim = c(8, 7)

named_clrs$cdr3 <- c(Short = "#482778", Medium = "#20938C", Long = "#C9E020")

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             clrs_specific = named_clrs$cdr3,
             # clrs_specific = named_colors$cdr3,
             title = "CDR3 Length", reduc = "rna.umap",
             meta_col = "cdr3_aa_length",
             plot_label = FALSE, legend_label = "CDR3 Length")
```

### Isotype

```{r bcr-isotype}
#| fig.dim = c(8, 7)

# uses colors defined in the alakazam package
plot_dimplot(seurat_obj = obj, data_source = data_desc,
             clrs_specific = named_colors$isotype,
             title = "Isotype", reduc = "rna.umap",
             meta_col = "isotype",
             plot_label = FALSE, legend_label = "Isotype")
```

### SHM frequency

You could also use a `FeaturePlot` to see the actual values.

```{r bcr-shm}
#| fig.dim = c(8, 7)

# note that the last bin's name is dependent on your dataset
# be careful with the order
plot_dimplot(seurat_obj = obj, data_source = data_desc,
             clrs_specific = named_colors$mu_freq_bins,
             title = "SHM Frequency", reduc = "rna.umap",
             meta_col = "mu_freq_bins", plot_label = FALSE,
             legend_label = "SHM Bins", sort_idents = FALSE,
             order = TRUE)
```

### V gene family

```{r bcr-vfamily}
#| fig.dim = c(8, 7)

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             title = "V Gene Family",
             clrs_specific = named_colors$v_call_family,
             meta_col = "v_call_family", plot_label = FALSE,
             legend_label = "V Gene Family", reduc = "rna.umap")
```

### Combining plots with patchwork

```{r bcr-combined}
#| fig.dim = c(14, 5)

p_iso <- plot_dimplot(obj, data_source = data_desc,
                      title = "Isotype", reduc = "rna.umap",
                      clrs_specific = named_colors$isotype,
                      meta_col = "isotype",
                      plot_label = FALSE, legend_label = "Isotype")
p_shm <- plot_dimplot(obj, data_source = data_desc,
                      title = "SHM", reduc = "rna.umap",
                      clrs_specific = named_colors$mu_freq_bins,
                      meta_col = "mu_freq_bins",
                      plot_label = FALSE, legend_label = "SHM Bins",
                      order = TRUE)
p_cdr3 <- plot_dimplot(obj, data_source = data_desc,
                       title = "CDR3 Length", reduc = "rna.umap",
                       clrs_specific = named_clrs$cdr3,
                       meta_col = "cdr3_aa_length",
                       plot_label = FALSE, legend_label = "CDR3 Length")

(p_iso + p_shm + p_cdr3) + plot_layout(nrow = 1)
```

---

## 3. ADT Expression

ADT features are stored as `adt_<marker>` in the Seurat object after
normalization. Because expression is continuous, these are best visualized
with Seurat's `FeaturePlot()` and a sequential color scale rather than
`plot_dimplot()`. `seurat_pipeline()` automatically normalizes the ADT assay
with CLR, so no extra step is needed.

ADT features are referenced as `"adt_<marker>"` (the assay key is prepended to
the feature name).

### Single marker

```{r adt-single}
#| fig.dim = c(8, 7)

# use CD19 as an example
FeaturePlot(obj, features = "adt_CD19",
            reduction = "rna.umap", order = TRUE, raster = FALSE) +
  scale_color_viridis_c(option = "G", direction = -1) + # name = "CD19"
  clean_dimplot2
```

### Multiple markers side by side

```{r adt-panel}
#| fig.dim = c(12, 8)

adt_markers <- rownames(GetAssayData(obj, assay = "ADT", layer = "data"))

FeaturePlot(obj, features = paste0("adt_", adt_markers),
            reduction = "rna.umap", order = TRUE, ncol = 3, raster = FALSE) &
  ggplot2::scale_color_viridis_c(option = "G", direction = -1) &
  clean_dimplot2
```

---

## 4. Different reductions

---

## Summary of key parameters

| Argument           | Purpose |
|--------------------|---------|
| `meta_col`         | Column in `@meta.data` to color by (default: `seurat_clusters`) |
| `title`            | Plot title (use for the data modality name) |
| `data_source`      | Plot subtitle (use for dataset / sample name) |
| `clrs_specific`    | Named color vector; auto-generated if omitted |
| `highlight`| Highlight a subset of levels over a grey background |
| `order`            | Draw cells with non-NA values on top |
| `plot_label`       | Toggle cluster labels |
| `reduc`            | Reduction to plot (default: `"umap"`) |
| `details`          | Optional extra subtitle line |
