---
title: "Main processing pipeline on real data: a single healthy control"
output:
  rmarkdown::html_vignette:
    code_folding: show
    df_print: kable
    fig_caption: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Main processing pipeline on real data: a single healthy control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 300, fig.align = "center",
                      out.width = "100%")
```

This vignette walks through the same `athanor` pipeline as `vignette("main_pipeline")`,
but on real single-cell data instead of simulated data:

1. **GEX processing**: `seurat_pipeline()` + `add_annotations()`
2. **BCR metadata**: adding AIRR-derived features to the Seurat object
3. **WNN**: `run_wnn()` integrates GEX and BCR embeddings via Seurat's Weighted Nearest
   Neighbors
4. **Concatenation**: `concatenate_gex_bcr()` combines gene expression and BCR
   embeddings or features into a single assay
5. **Evaluation**: `calc_neighbor_matches()` and `calc_adt_correlation()`
   check how well each modality's neighbor graph tracks real BCR features
   and real surface protein levels

The data are a subsample of one healthy (non-diabetic) control from
[GSE270142](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270142)
([Nicholas et al.](https://www.sciencedirect.com/science/article/pii/S2211124725001962)):
1,500 PBMCs with paired 10x gene expression, ADT (surface protein), and BCR
sequencing. BCR embeddings (AntiBERTa2) and AIRR annotations (mutation
frequency, isotype, CDR3 features, V/J calls) come from real repertoire
processing, not simulation. The fixture is bundled with the package under
`inst/extdata/diabetes_nicholas_108_ND/`.

---

## Setup

```{r libs}
library(athanor)
library(dplyr)
library(ggplot2)
library(Matrix)
library(patchwork)
library(Seurat)

set.seed(42)

data_desc <- "Diabetes Nicholas (108_ND, healthy control)"
path_data <- system.file("extdata", "diabetes_nicholas_108_ND",
                         package = "athanor")
```

---

## 1. GEX processing

### Build the initial object

We load the real RNA and ADT count matrices and attach them to a single
Seurat object. `cell_id` (`"<sample>_<barcode>"`) is required downstream by
`run_wnn()`; here it is already baked into the column names of the bundled
matrices.

```{r load-gex}
rna_counts <- readRDS(file.path(path_data, "rna_counts.rds"))
adt_counts <- readRDS(file.path(path_data, "adt_counts.rds"))

obj <- CreateSeuratObject(counts = rna_counts, assay = "RNA")
obj[["ADT"]] <- CreateAssay5Object(counts = adt_counts)

obj$cell_id <- Cells(obj)
obj$percent.mt <- PercentageFeatureSet(obj, pattern = "^MT-")

# real cells, real genes
dim(obj)
head(obj$cell_id)
```

### Run the GEX pipeline

`seurat_pipeline()` runs QC filtering, normalization, variable feature
selection, scaling, PCA, neighbor finding, and UMAP. Unlike the simulated
vignette, we now filter on real QC metrics (`nfeatures_RNA`, `perc_mt`).

```{r gex-pipeline}
num_pcs <- 50
num_dims <- 20
k <- 20

obj <- seurat_pipeline(seurat_obj = obj, nfeatures_RNA = 200, perc_mt = 15,
                       num_pcs = num_pcs, num_dims = num_dims, k_param = k,
                       cluster_res = 0.5, verbose = FALSE)

# post-QC cell count, reductions, and clusters produced
dim(obj)
names(obj@reductions)
levels(obj$seurat_clusters)
```

```{r gex-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj, plot_title = "GEX", data_source = data_desc,
             meta_col = "seurat_clusters", reduc = "rna.umap")
```

### Annotate clusters

`add_annotations()` maps a data frame of cluster-to-cell-type assignments
onto the Seurat object. Rather than simulating labels, we call each cluster
from real marker gene expression: `IGHD`/`TCL1A` mark naive B cells,
`CD27`/`TNFRSF13B` mark memory B cells, `MZB1`/`XBP1` mark plasma cells, and
`CD14`/`LYZ` (monocytes) flag non-B contaminants that inevitably survive
FACS/10x sorting.

```{r marker-dotplot}
#| fig.dim = c(9, 5)

markers <- c("IGHD", "TCL1A", "CD19", "MS4A1", "CD27", "TNFRSF13B", "MZB1",
             "XBP1", "CD3D", "CD3E", "CD8A", "CD14", "LYZ")
DotPlot(obj, features = markers, group.by = "seurat_clusters") + RotatedAxis()
```

Cluster 5 is clearly non-B cells (monocytes) rather than a failure of the sort - real PBMC preparations
always carry some contaminants. We drop them before annotating the
remaining, genuine B cell clusters.

```{r drop-non-b}
b_clusters <- c("0", "1", "2", "3", "4", "6", "7")
obj <- subset(obj, subset = seurat_clusters %in% b_clusters)
obj$seurat_clusters <- droplevels(obj$seurat_clusters)

dim(obj)
```

```{r annotate}
# read off from the dot plot above (naive: high IGHD/TCL1A, low CD27;
# memory: low IGHD/TCL1A, high CD27/TNFRSF13B; plasma: high MZB1/XBP1,
# low MS4A1). Clusters, in order: 0, 1, 2, 3, 4, 6, 7
cell_types <- c("Naive B cells", "Memory B cells", "Naive B cells",
                "Memory B cells", "Plasma cells", "Naive B cells",
                "Plasma cells")

annotations_df <- data.frame(CellType = cell_types)

obj <- add_annotations(seurat_obj = obj, annotations_df = annotations_df,
                       cell_types_col = "CellType",
                       clusters_col = "seurat_clusters",
                       annotations_col = "annotated_clusters")

table(obj$annotated_clusters)
```

```{r annotated-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "GEX", reduc = "rna.umap",
             meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

---

## 2. BCR data

### Metadata

`gex_add_airr()` adds AIRR-derived features (isotype, mutation frequency,
CDR3 length, V/J gene calls, chain pairing, ...) onto a Seurat object. This
is real repertoire data (from an nf-core/airrflow + Immcantation pipeline),
not simulated.

```{r load-airr}
combined_airr <- readRDS(file.path(path_data, "combined_airr.rds"))

colnames(combined_airr)
```

```{r bcr-meta}
new_cols <- c("cell_id", "locus", "v_call", "v_call_family", "v_call_gene",
             "d_call", "d_call_family", "d_call_gene", "j_call",
             "j_call_family", "j_call_gene", "c_call", "isotype", "mu_freq",
             "clone_id", "cdr3_aa_length")

obj <- gex_add_airr(seurat_obj = obj, airr_type = "BCR",
                    combined_airr = combined_airr, new_cols = new_cols)

obj[[]] %>%
  select(mu_freq, cdr3_aa_length, isotype) %>%
  summary()
```

### Embeddings

We load real AntiBERTa2 embeddings (heavy + light chain, concatenated) for
this subject. Not every GEX cell has a paired BCR sequence, so the embedding
matrix has fewer columns than `obj`. Passing `combined_airr` through to
`bcr_embeddings_pipeline()` adds the same AIRR columns (including `Has_BCR`
and `paired_light`, needed by `merge_gex_bcr()` below) onto the BCR object.

```{r load-bcr-embeddings}
embeddings <- readRDS(file.path(path_data, "bcr_embeddings.rds"))

# restrict to the B cells that survived filtering above
embeddings <- embeddings[, intersect(colnames(embeddings), Cells(obj))]

# dimensions by cells; fewer cells than obj since not all have paired BCR data
dim(embeddings)
embeddings[1:5, 1:5]
```

```{r bcr-pipeline}
embeddings_obj <- bcr_embeddings_pipeline(embeddings = embeddings,
                                          embedding_type = "AntiBERTa2",
                                          combined_airr = combined_airr,
                                          new_cols = new_cols,
                                          num_pcs = num_pcs,
                                          num_dims = num_dims, k_param = k,
                                          verbose = TRUE)
```

Rather than color by cell type again (already shown for GEX above), we check
whether the embedding space actually preserves real BCR features: isotype
and somatic hypermutation (SHM) frequency. If AntiBERTa2 embeddings are
biologically meaningful, cells should organize by these labels even though
the model never saw them during training.

```{r bcr-umap}
#| fig.dim = c(14, 6)

embeddings_obj <- bin_mu_freq(embeddings_obj)

p_isotype <- plot_dimplot(seurat_obj = embeddings_obj, data_source = data_desc,
                          clrs_specific = named_colors$isotype,
                          plot_title = "BCR Embeddings", reduc = "bcr.umap",
                          meta_col = "isotype", plot_label = FALSE,
                          legend_label = "Isotype")
p_mu_freq <- plot_dimplot(seurat_obj = embeddings_obj, data_source = data_desc,
                          clrs_specific = named_colors$mu_freq_bins,
                          plot_title = "BCR Embeddings", reduc = "bcr.umap",
                          meta_col = "mu_freq_bins", plot_label = FALSE,
                          legend_label = "SHM Frequency")

p_isotype + p_mu_freq
```

---

## 3. WNN (Embeddings)

`run_wnn()` integrates two modalities via Weighted Nearest Neighbors. It
expects:

- a Seurat object with a `cell_id` column in the metadata
- a BCR `embeddings` matrix (features x cells) whose column names match
  `cell_id`, covering *exactly* the cells in the Seurat object (unlike
  `merge_gex_bcr()`, it does not subset the GEX side down to shared cells)

Since real cells don't all have paired BCR data, we restrict to the subset
that does before calling `run_wnn()`.

```{r run-wnn}
obj_bcr <- subset(obj, cell_id %in% colnames(embeddings))

obj_wnn <- run_wnn(seurat_obj = obj_bcr, embeddings = embeddings,
                   embedding_type = "AntiBERTa2",
                   pc_gex = 10, pc_bcr = 10, k_param = 20, cluster = TRUE,
                   cluster_res = list("GEX" = 0.5, "BCR" = 0.5, "WNN" = 0.5),
                   verbose = FALSE)

# bin the mutation frequency for plotting
obj_wnn <- bin_mu_freq(obj_wnn)

names(obj_wnn@reductions)
```

### Visualization

Cell type is already shown for GEX and BCR embeddings above, so here we lay
the three modalities side by side on a variable neither one was built to
predict: class-switch status (`isotype_stage`, derived from the real
isotype call). Agreement across GEX/BCR/WNN is a sanity check that the
modalities are capturing consistent biology rather than each inventing its
own structure.

```{r wnn-umap}
#| fig.dim = c(14, 5)

p_gex <- plot_dimplot(obj_wnn, plot_title = "GEX", data_source = data_desc,
                      clrs_specific = named_colors$isotype_stage,
                      meta_col = "isotype_stage", reduc = "rna.umap",
                      plot_label = FALSE, legend_label = "Switch Status")
p_bcr <- plot_dimplot(obj_wnn, plot_title = "BCR", data_source = data_desc,
                      clrs_specific = named_colors$isotype_stage,
                      meta_col = "isotype_stage", reduc = "bcr.umap",
                      plot_label = FALSE, legend_label = "Switch Status")
p_wnn <- plot_dimplot(obj_wnn, plot_title = "WNN", data_source = data_desc,
                      clrs_specific = named_colors$isotype_stage,
                      meta_col = "isotype_stage", reduc = "wnn.umap",
                      plot_label = FALSE, legend_label = "Switch Status")

(p_gex + p_bcr + p_wnn) + plot_layout(nrow = 1, guides = "collect")
```

The WNN embedding is still worth seeing colored by cell type once, since
it's the first (and only) time these B cell subtype calls are shown against
the *jointly* integrated space rather than a single modality:

```{r wnn-umap-celltype}
#| fig.dim = c(8, 6)

plot_dimplot(obj_wnn, plot_title = "WNN", data_source = data_desc,
            clrs_specific = named_colors$cell_types_celltypist,
            meta_col = "annotated_clusters", reduc = "wnn.umap",
            plot_label = FALSE, legend_label = "Cell Type")
```

Modality weights:

```{r wnn-mw-box}
#| fig.dim = c(14, 5)

# by isotype
plot_mws(seurat_obj = obj_wnn, second_assay = "BCR",
         clrs_specific = named_colors$isotype, split_by = "isotype",
         facet_by = "annotated_clusters", y_axis_label = "Isotype")

# by SHM frequency
plot_mws(seurat_obj = obj_wnn, second_assay = "BCR",
         clrs_specific = named_colors$mu_freq_bins, split_by = "mu_freq_bins",
         facet_by = "annotated_clusters", y_axis_label = "SHM Frequency Bins")
```

`object_overview()` prints a brief description of the post-WNN object,
including assay sizes and the number of PCs used per modality.

```{r wnn-summary}
object_overview(obj_wnn)
```

---

## 4. Concatenation (Embeddings)

`concatenate_gex_bcr()` combines the RNA counts with processed BCR
embeddings into a new `RNA_BCR` assay, then runs PCA and UMAP on the joint
space.

```{r concatenate-embeddings}
# merge the BCR embeddings into the GEX obj
gex_bcr_obj <- merge_gex_bcr(gex_obj = obj, bcr_obj = embeddings_obj,
                             transfer_reductions = TRUE, verbose = TRUE)

obj_cat <- concatenate_gex_bcr(seurat_obj = gex_bcr_obj, stage = "embed",
                               input_type = "embeddings",
                               gex_reduction = "rpca",
                               normalize = TRUE, num_dims = num_dims)

names(obj_cat@assays)
names(obj_cat@reductions)
```

```{r concatenated-embeddings-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj_cat, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "Concatenated GEX & BCR Embeddings",
             reduc = "rna_bcr.umap", meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

```{r concatenated_embeddings-summary}
object_overview(obj_cat)
```

---

## 5. Concatenation (Features)

`concatenate_gex_bcr()` can also combine the RNA counts with processed BCR
feature columns (rather than embeddings) into a new `RNA_BCR` assay.

```{r concatenate-features}
obj_cat_feat <- concatenate_gex_bcr(seurat_obj = obj, stage = "reduced",
                                    cols_to_include =
                                      c("cdr3_aa_length", "isotype", "mu_freq"),
                                    normalize = TRUE, num_dims = num_dims)

names(obj_cat_feat@assays)
names(obj_cat_feat@reductions)
```

```{r concatenated-features-umap}
#| fig.dim = c(8, 6)

plot_dimplot(seurat_obj = obj_cat_feat, data_source = data_desc,
             clrs_specific = named_colors$cell_types_celltypist,
             plot_title = "Concatenated GEX & BCR Features", reduc = "rna_bcr.umap",
             meta_col = "annotated_clusters",
             plot_label = FALSE, legend_label = "Cell Type")
```

```{r concatenated-features-summary}
object_overview(obj_cat_feat)
```

---

## 6. Evaluation

Two ways to check whether these joint representations actually capture real
biology, rather than just concatenating noise, following the same approach
used in the `athanor` manuscript: do a cell's neighbors share its real BCR
features, and do they share its real surface protein (ADT) levels? Both
checks need real data - simulated features/embeddings would validate the
pipeline's plumbing but not whether the embeddings themselves are
informative. We compare across all five representations built above: GEX,
BCR, WNN, Concatenated Embeddings, and Concatenated Features.

We first regenerate a `RNA.nn` neighbor object on `obj_wnn` (dropped when
`obj` was subset earlier - `FindNeighbors()` neighbor objects don't survive
subsetting) so GEX can be compared alongside BCR and WNN.

```{r regen-rna-nn}
obj_wnn <- regen_reduc(obj_wnn, pca_name = "rpca", assay = "RNA",
                       num_dims = 10, k_param = 20, verbose = FALSE)

names(obj_wnn@neighbors)
```

```{r approach-setup, include = FALSE}
approach_levels <- c("GEX", "BCR", "WNN", "Concatenated\nEmbeddings",
                     "Concatenated\nFeatures")
approach_colors <- c("GEX" = named_colors$datatype[["GEX"]],
                     "BCR" = named_colors$datatype[["BCR"]],
                     "WNN" = named_colors$datatype[["WNN"]],
                     "Concatenated\nEmbeddings" =
                       named_colors$datatype[["GEX_BCR"]],
                     "Concatenated\nFeatures" = "#14506b")
feature_labels <- c(annotated_clusters_bcr = "BCR-Based\nCell Type",
                   cdr3_aa_length = "CDR3\nLength",
                   isotype_stage = "Isotype\nStage",
                   locus_light = "Light Chain\nLocus",
                   mu_freq_bins_binary = "SHM\nFrequency",
                   v_call_family = "V Gene\nFamily")
```

### Neighbor matches

`calc_neighbor_matches()` asks, for each cell in each neighbor graph, what
fraction of its k nearest neighbors share the same real BCR feature (e.g.
isotype, CDR3 length, V gene family). Higher scores mean the graph's notion
of "similar cells" agrees with real repertoire biology, not just
transcriptome-driven clustering. `return_mean = FALSE` keeps one score per
cell so we can see the full distribution, not just an average.

```{r neighbor-matches}
nn_scores <-
  bind_rows(
    calc_neighbor_matches(obj_wnn, return_mean = FALSE, verbose = FALSE) %>%
      mutate(Approach = recode(Graph, "RNA.nn" = "GEX", "BCR.nn" = "BCR",
                               "w.nn" = "WNN")),
    calc_neighbor_matches(obj_cat, nn_names = "RNA_BCR.nn",
                         return_mean = FALSE, verbose = FALSE) %>%
      mutate(Approach = "Concatenated\nEmbeddings"),
    calc_neighbor_matches(obj_cat_feat, nn_names = "RNA_BCR.nn",
                         return_mean = FALSE, verbose = FALSE) %>%
      mutate(Approach = "Concatenated\nFeatures")
  ) %>%
  mutate(Approach = factor(Approach, levels = approach_levels))

nn_scores %>%
  group_by(Approach, Feature) %>%
  summarize(Score = mean(Score), .groups = "drop")
```

```{r neighbor-matches-plot}
#| fig.dim = c(16, 6)

ggplot(nn_scores, aes(x = Approach, y = k * Score, fill = Approach)) +
  geom_violin(alpha = 0.5, color = "grey50") +
  geom_boxplot(outlier.shape = NA, width = 0.3) +
  scale_fill_manual(values = approach_colors) +
  scale_y_continuous(breaks = seq(0, k, 4),
                    sec.axis =
                      sec_axis(transform = ~ . / k,
                              name = "Fraction of k nearest neighbors that match",
                              breaks = seq(0, 1, 0.25),
                              labels = scales::label_percent())) +
  facet_grid(cols = vars(Feature),
            labeller = labeller(Feature = feature_labels)) +
  labs(x = NULL, y = "# of k Nearest Neighbors with Matching Label",
      plot_title = "Neighbor matching scores", subtitle = data_desc) +
  theme_bw_custom + labels_rotate_x + labels_standard +
  theme(legend.position = "none")
```

### ADT correlation

`calc_adt_correlation()` correlates each cell's real surface protein (ADT)
level with the mean level among its neighbors, for markers relevant to B
cell subtyping: IgM/IgD (naive vs. class-switched), CD21 (marginal
zone/memory), and CD27 (memory). A modality whose neighbor graph tracks true
surface phenotype should show higher correlations here.

```{r adt-correlation}
adt_features <- c("IgM", "IgD", "CD21", "CD27.1")

adt_scores <-
  bind_rows(
    calc_adt_correlation(obj_wnn, features_adt = adt_features,
                        verbose = FALSE) %>%
      mutate(Approach = recode(Graph, "RNA.nn" = "GEX", "BCR.nn" = "BCR",
                               "w.nn" = "WNN")),
    calc_adt_correlation(obj_cat, features_adt = adt_features,
                        verbose = FALSE) %>%
      filter(Graph == "RNA_BCR.nn") %>%
      mutate(Approach = "Concatenated\nEmbeddings"),
    calc_adt_correlation(obj_cat_feat, features_adt = adt_features,
                        verbose = FALSE) %>%
      mutate(Approach = "Concatenated\nFeatures")
  ) %>%
  mutate(Approach = factor(Approach, levels = approach_levels))

adt_scores
```

```{r adt-correlation-plot}
#| fig.dim = c(16, 5)

ggplot(adt_scores, aes(x = Approach, y = Score, fill = Approach)) +
  geom_point(size = 3, color = "grey25", pch = 21) +
  scale_fill_manual(values = approach_colors) +
  scale_y_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1)) +
  facet_grid(cols = vars(Feature)) +
  labs(x = NULL, y = "Spearman Correlation (cell vs. neighbor mean)",
      plot_title = "ADT correlation", subtitle = data_desc) +
  theme_bw_custom + labels_rotate_x + labels_standard +
  theme(legend.position = "none")
```

---

## Summary of the pipeline

| Step | Function | Key output(s) |
|------|----------|-----------|
| Load real 10x GEX/ADT counts | `CreateSeuratObject()` | Seurat object with `cell_id` |
| GEX pipeline | `seurat_pipeline()` | `RNA` assay, `rpca`, `rna.umap`, `seurat_clusters` |
| Load real BCR embeddings | `readRDS()` | Matrix of AntiBERTa2 embedding values |
| BCR pipeline | `bcr_embeddings_pipeline()` | `BCR` assay, `bpca`, `bcr.umap` |
| Annotation | `add_annotations()` | `annotated_clusters` in metadata |
| Addition of BCR features | `gex_add_airr()` | BCR columns in object metadata |
| Weighted nearest neighbors | `run_wnn()` | `wnn.umap` |
| Concatenation | `concatenate_gex_bcr()` | `RNA_BCR` assay, `rna_bcr.pca`, `rna_bcr.umap` |
| Description | `object_overview()` | printed summary |
| Neighbor matching | `calc_neighbor_matches()` | fraction of neighbors sharing real BCR features |
| ADT correlation | `calc_adt_correlation()` | cell vs. neighbor-mean correlation for real surface markers |
